Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-01-10 analysis/paper/article.qmd
Minor formatting edits.

Attend any workshop, conference, or panel that includes early career researchers and the conversation will be steered inexorably towards the academic job market: who is hiring, who has attained a tenure-track position, and who is out of luck this season. This obsessive focus on career trajectories is warranted, particularly within the discipline of anthropology. Between 1985 and 2014, the number of PhDs produced by American anthropology departments increased by 70% (**Speakman et al., 2018a**), but the number of tenure-track positions advertised has stagnated or declined*.* As of2019, 53.5% of colleges and universities had replaced tenure-eligible lines with contingent positions (**AAUP 2022**); today, 71% of faculty in the US are non-tenure-track(**Culver and Kezar 2022***).* The erosion of permanent jobs in American higher education is linked to a complex intersection of political and economic factors (**Beck, under review**), including decreasing federal support for higher education, concomitant increases in student debt (**Gusterson 2017**), a pronounced shift in university investment from faculty to administration (Graeber 2018: 162–163), and the growing privatization and market orientation of scientific research (**Mirowski 2011**). Whatever the cause, research within anthropology has consistently shown that at most 20% of PhDs in the discipline are obtaining jobs in American anthropology departments (**Speakman et al. 2018b**), while the top-ranking programmes are placing fewer than one in three graduates in tenure-track jobs (**Mackie and Rockwell 2023**).

These cultural and economic shifts are also well encapsulated by anecdata; any conversation with a junior scholar will reveal the impact of these larger professional transformations on personal and professional trajectories. The first author on this paper has applied to five academic positions over the course of his career, while the last author has applied to one hundred. A twenty-fold increase in the number of applications required to obtain a workable long-term academic position may seem preposterous, but these changing requirements for applicants are familiar to anyone who has obtained their PhDs since the Great Recession. This competitive landscape may have deeper roots than the last economic crisis; writing more than three decades ago, Rabinow pinpointed “an awareness… that standards have changed during the last thirty years and the quantitative and qualitative demands for entry into the system are immeasurably higher now” (**1992: 66**).

@@

We argue that to understand the dynamics of hiring in academic archaeology, we must start at the beginning, with the job ad itself. There is an abundance of recent work on who does and does not get hired into tenure-track positions in anthropology and archaeology (Kawa 2019; Mackie and Rockwell 2023), but less scholarship on how those hires unfold. Research on hiring dynamics has amply demonstrated that *where* you get your PhD has an enormous impact on if and where you get a tenure-track position (Kawa et al. 2019; Mackie and Rockwell 2023), but how about *what* you study? Are there topical, methodological, or regional specialties that are in consistent demand, or is there continual flux generated by intellectual fads, economic incentives, and political attention? Are job applications more demanding now than they were in the past, or is this a specious perception generated by the sheer volume of applications necessary to obtain a long-term academic position?

@@

The academic job market is a source of uncertainty, for both job seekers and hiring committees. Job seekers must search for ads that are well-matched to the skill sets and professional experiences that they have taken years to cultivate. Hiring committeesmust sift through tens to hundreds of applicants in the hope of finding a candidate that meets all of their needs and helps to realize<!--We should go with American spelling, no? With your Australian inclinations and my growing incorporation of Irish orthography into my writing, this paper is gonna be a mess...--> their visions of their program and the discipline. Finding the right candidate for the position—or the right position for the candidate—is akin to finding a needle in a haystack.

@@

In anthropology in the US the number of doctoral anthropology graduates has increased by about 70% over the past 30 years, but the number of new faculty positions has not increased proportionally [@speakman2018choosing]. New faculty positions have dwindled, in part due to the removal of the of the Age Discrimination in Employment Act (ADEA) exemption in 1994 which prohibited mandatory retirement ages in higher education (**Earle and Delpo Kulow 2015**). When combined with the institution of tenure in American academia, the ADEA exemption allowed [faculty to stay in their posts for as long as they liked] the median age for faculty in the U.S. now ranks among the highest for all professions (**Citation**). In tandem with the gradual de-investment in American higher education since the **XXX (Mirowski**) and the aftershocks of the 2008 recession, the ADEA exemption has cultivated an environment where new lines are few and far between. Among biological anthropologists, @Passalacqua_2018 found a ratio of 0.81 PhDs to job academic advertisements per year, concluding that academic positions in biological anthropology are barely at sustainable levels. This echoes findings from other fields. In biomedical sciences, for example, there is one tenure-track position in the US for approximately every 6.3 PhD graduates [@Ghaffarzadegan_2015]. **Citation of Larson et al. 2014**.

>>>>

Attend any workshop, conference, or panel that includes early career researchers and the conversation will be steered inexorably towards the academic job market: who is hiring, who has attained a tenure-track position, and who is out of luck this season. This obsessive focus on career trajectories is warranted, particularly within the discipline of anthropology. Between 1985 and 2014, the number of PhDs produced by American anthropology departments increased by 70% (**Speakman et al., 2018a**), but the number of tenure-track positions advertised has stagnated or declined*.* As of2019, 53.5% of colleges and universities had replaced tenure-eligible lines with contingent positions (**AAUP 2022**); today, 71% of faculty in the US are non-tenure-track(**Culver and Kezar 2022***).* The erosion of permanent jobs in American higher education is linked to a complex intersection of political and economic factors (**Beck, under review**), including decreasing federal support for higher education, concomitant increases in student debt (**Gusterson 2017**), a pronounced shift in university investment from faculty to administration (**Graeber 2018: 162–163**), and the growing privatization and market orientation of scientific research (**Mirowski 2011**). Whatever the cause, research within anthropology has consistently shown that at most 20% of PhDs in the discipline are obtaining jobs in American anthropology departments (**Speakman et al. 2018b**), while the top-ranking programmes are placing fewer than one in three graduates in tenure-track jobs (**Mackie and Rockwell 2023**).

These cultural and economic shifts are also well encapsulated by anecdata; any conversation with a junior scholar will reveal the impact of these larger professional transformations on personal and professional trajectories. The first author on this paper has applied to five academic positions over the course of his career, while the last author has applied to one hundred. A twenty-fold increase in the number of applications required to obtain a workable long-term academic position may seem preposterous, but these changing requirements for applicants are familiar to anyone who has obtained their PhDs since the Great Recession. This competitive landscape has deeper roots than the last economic crisis; writing more than three decades ago, Rabinow pinpointed “an awareness… that standards have changed during the last thirty years and the quantitative and qualitative demands for entry into the system are immeasurably higher now” (**1992: 66**).

@@

We argue that to understand the dynamics of hiring in academic archaeology, we must start at the beginning, with the job ad itself. There is an abundance of recent work on who does and does not get hired into tenure-track positions in anthropology and archaeology (**Kawa 2019; Mackie and Rockwell 2023**), but less scholarship on how those hires unfold. Research on hiring dynamics has amply demonstrated that *where* you get your PhD has an enormous impact on if and where you get a tenure-track position (**Kawa et al. 2019; Mackie and Rockwell 2023**), but how about *what* you study? Are there topical, methodological, or regional specialties that are in consistent demand, or is there continual flux generated by intellectual fads, economic incentives, and political attention? Are job applications more demanding now than they were in the past, or is this a specious perception generated by the sheer volume of applications necessary to obtain a long-term academic position?

@@

The academic job market is a source of uncertainty, for both job seekers and hiring committees. Job seekers must search for ads that are well-matched to the skill sets and professional experiences that they have taken years to cultivate. Hiring committeesmust sift through tens to hundreds of applicants in the hope of finding a candidate that meets all of their needs and helps to realize their visions of their program and the discipline. Finding the right candidate for the position—or the right position for the candidate—is akin to finding a needle in a haystack.

@@

In anthropology in the US the number of doctoral anthropology graduates has increased by about 70% over the past 30 years, but the number of new faculty positions has not increased proportionally [@speakman2018choosing]. New faculty positions have dwindled, in part due to the removal of the of the Age Discrimination in Employment Act (ADEA) exemption in 1994 which prohibited mandatory retirement ages in higher education (**Earle and Delpo Kulow 2015**). When combined with the institution of tenure in American academia, the ADEA exemption allowed [faculty to stay in their posts for as long as they liked] the median age for faculty in the U.S. now ranks among the highest for all professions (**Citation**). In tandem with the gradual de-investment in American higher education since the **XXX (Mirowski 2011**) and the aftershocks of the 2008 recession, the ADEA exemption has cultivated an environment where new lines are few and far between. Among biological anthropologists, @Passalacqua_2018 found a ratio of 0.81 PhDs to job academic advertisements per year, concluding that academic positions in biological anthropology are barely at sustainable levels. This echoes findings from other fields. In biomedical sciences, for example, there is one tenure-track position in the US for approximately every 6.3 PhD graduates [@Ghaffarzadegan_2015]. **Citation of Larson et al. 2014**.
